Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of neutron star - black hole binary observations

### 2017-01-18 Paper/journal_submission/paper.tex
final changes before resubmission

Gravitational waves emitted by coalescing NSBH binaries carry subtle hints of
the NS EoS from inspiral through to merger. During early inspiral, the tidal
dephasing is relatively weak and has a frequency dependence equivalent to a
$5^{th}$ Post-Newtonian (PN) order effect~\cite{Vines2011}. Closer to merger,
a disruptive fate of the NS
can result in a strong suppression of GW emission above a cut-off 
frequency~\cite{Pannarale:2015jia}. Some past studies of tidal measurements
with NSBH binaries have used PN inspiral-only waveforms~\cite{Maselli:2013rza}.
In doing so, however, they ignore (i) the merger signal which could contain significant
information for NSBHs, and (ii) the errors due to unknown vaccum terms in PN 
waveforms, which could dominate over the tidal terms themselves~\cite{Barkett2015,
Yagi:2014}.
% 
Some other studies that account for merger effects via the use of complete
numerical simulations~\cite{Foucart:2013psa}, are limited in the binary
parameter space they sample.
% 
Others, that do the same through the use of phenomenological waveform
models~\cite{Lackey2011,Lackey:2013axa} use the Fisher matrix to estimate
$\lambdans$ measurement errors. Fisher matrix estimates may become
unreliable at realistic signal-to-noise ratios (SNR)~\cite{Vallisneri:2007ev},
such as those as we might expect in the upcoming observing runs of GW
detectors~\cite{Abadie:2010cf}, and we supplement such studies with a
fully Bayesian treatment of the problem here.

@@

% ###########################################################
% ###########################################################
\section{Choice of underlying BBH model in LEA}
% ###########################################################
% 
\begin{figure}
\centering    
\includegraphics[width=1.05\columnwidth]{match-q-spin1-PhenomC-mf0_01.pdf}
\caption{%
We compare two alternatives of the tidal NSBH model from Ref.~\cite{
Lackey:2013axa}, which differ in their underlying BBH prescriptions. One which
we use in this study uses SEOBNRv2, while the other uses IMRPhenomC as its
base. In this figure, we show the normalized overlap (match) between the two for
$2,000,000$ points sampled uniformly in the NSBH parameter space. We find that
for $q\gtrsim 4$ the discrepancies between IMRPhenomC and SEOBNRv2 as reported
in~\cite{Kumar:2015tha} dominate over tidal terms.
\label{fig:PhenomC_vs_SEOBNRv2_LEA}
}
\end{figure}
%
The waveform model used in this paper is a variant of those calibrated
in Ref~\cite{Lackey:2013axa}. In that work, the authors also calibrate
a tidal prescription with the phenomenological model IMRPhenomC~\cite{
Santamaria:2010yb}
as the base BBH model. Previous work~\cite{Kumar:2015tha} has shown that 
IMRPhenomC can exhibit pathological behavior for mass-ratios ever so
$q\gtrsim 4$ and/or non-zero black hole spins. We compute noise-weighted
inner-products between the two variants for $2,000,000$ points sampled
over the NSBH parameter space, and show the results in Fig.~\ref{fig:PhenomC_vs_SEOBNRv2_LEA}. We restrict the comparison to frequencies
that are affected by the tidal disruption of the NS, by integrating
the inner-products from $f = \mathrm{max}(15, 0.01/M)$~Hz
(where $M$ is expressed in seconds ($1M_\odot \simeq 4.925\mu$S, see
Eq.~(32-34) of~\cite{Lackey:2013axa}).
% 
We find that the differences between the two BBH approximants dominate
over the tidal terms that are added to their phasing. As SEOBNRv2 is
more reliable than IMRPhenomC, we recommend the use of SEOBNRv2-based
LEA+ in upcoming LIGO-Virgo analyses.

>>>>

Gravitational waves emitted by coalescing NSBH binaries carry subtle hints of
the NS EoS from inspiral through to merger. During early inspiral, the tidal
dephasing is relatively weak and has a frequency dependence equivalent to a
$5^{th}$ Post-Newtonian (PN) order effect~\cite{Vines2011}. Closer to merger,
a disruptive fate of the NS
can result in a strong suppression of GW emission above a cut-off 
frequency~\cite{Pannarale:2015jia}. Some past studies of tidal measurements
with NSBH binaries have used PN inspiral-only waveforms~\cite{Maselli:2013rza}.
In doing so, however, they ignore (i) the merger signal which could contain significant
information for NSBHs, and (ii) the errors due to unknown vaccum terms in PN 
waveforms, which could dominate over the tidal terms themselves~\cite{Barkett2015,
Yagi:2014}.
% 
Some other studies that account for merger effects via the use of complete
numerical simulations~\cite{Foucart:2013psa}, are limited in the binary
parameter space they sample.
% 
Others, that do the same through the use of phenomenological waveform
models~\cite{Lackey2011,Lackey:2013axa} use the Fisher matrix to estimate
$\lambdans$ measurement errors. Fisher matrix estimates may become
unreliable at realistic signal-to-noise ratios (SNR)~\cite{Vallisneri:2007ev},
such as those as we might expect in the upcoming observing runs of GW
detectors~\cite{Abadie:2010cf}, and we improve such studies with a
fully Bayesian treatment of the problem here.

@@

% ###########################################################
% ###########################################################
\section{Choice of underlying BBH model in LEA}
% ###########################################################
% 
\begin{figure}
\centering    
\includegraphics[width=1.05\columnwidth]{match-q-spin1-PhenomC-mf0_01.pdf}
\caption{%
We compare two alternatives of the tidal NSBH model from Ref.~\cite{
Lackey:2013axa}, which differ in their underlying BBH prescriptions. One which
we use in this study uses SEOBNRv2, while the other uses IMRPhenomC as its
base. In this figure, we show the normalized overlap (match) between the two for
$2,000,000$ points sampled uniformly in the NSBH parameter space. We find that
for $q\gtrsim 4$ the discrepancies between IMRPhenomC and SEOBNRv2 as reported
in~\cite{Kumar:2015tha} dominate over tidal terms.
\label{fig:PhenomC_vs_SEOBNRv2_LEA}
}
\end{figure}
%
The waveform model used in this paper is a variant of those calibrated
in Ref~\cite{Lackey:2013axa}. In that work, the authors also calibrate
a tidal prescription with the phenomenological model IMRPhenomC~\cite{
Santamaria:2010yb}
as the base BBH model. Previous work~\cite{Kumar:2015tha,Kumar:2016dhh} has 
shown that IMRPhenomC can exhibit pathological behavior for mass-ratios
$q\gtrsim 4$ and/or non-zero black hole spins. We compute noise-weighted
inner-products between the two variants for $2,000,000$ points sampled
over the NSBH parameter space, and show the results in 
Fig.~\ref{fig:PhenomC_vs_SEOBNRv2_LEA}. We restrict the comparison to frequencies
that are affected by the tidal disruption of the NS, by integrating
the inner-products from $f = \mathrm{max}(15, 0.01/M)$~Hz
(where $M$ is expressed in seconds ($1M_\odot \simeq 4.925\mu$S, see
Eq.~(32-34) of~\cite{Lackey:2013axa}).
% 
We find that the differences between the two BBH approximants cause mismatches
of a few percent, while the tidal corrections contribute at a sub-percent
level. We conclude that the differences of the underlying BBH model in LEA+
dominates over their tidal calibration, and since SEOBNRv2 has been shown to
be more reliable than IMRPhenomC~\cite{Kumar:2015tha,Kumar:2016dhh}, we 
recommend the use of SEOBNRv2-based LEA+ in upcoming LIGO-Virgo analyses.
